### 2025-07-14 paper/quarto_aaai/paper.tex
one more render after updated quarto template

\iffalse
%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \title{My Publication Title --- Multiple Authors}
\author {
    % Authors
    First Author Name\textsuperscript{\rm 1},
    Second Author Name\textsuperscript{\rm 2},
    Third Author Name\textsuperscript{\rm 1}
}
\affiliations {
    % Affiliations
    \textsuperscript{\rm 1}Affiliation 1\\
    \textsuperscript{\rm 2}Affiliation 2\\
    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
}
\fi
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage{placeins}
\usepackage{siunitx}
\sisetup{uncertainty-mode = separate}
\DeclareMathSizes{10}{9}{7}{6.5}

@@

Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
training regime explicitly geared towards improving the explanatory
capacity of models that, in high-level terms, we define as the extent to
which valid explanations derived for a model can be deemed plausible
with respect to the underlying data and global actionability constraints
(we refine this notion in Def. \ref{def-explainability}). For
simplicity, we refer to models with high explanatory capacity as
\emph{explainable}. To the best of our knowledge, Counterfactual
Training represents the first attempt to achieve more explainable models
by employing counterfactual explanations already in the training phase.

@@

In the context of counterfactual explanations, plausibility has broadly
been defined as the degree to which generated CEs comply with the
underlying data-generating process
\citep{altmeyer2024faithful, guidotti2022counterfactual, poyiadzi2020face}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. An AR recommendations must also be actionable, i.e.,
possible to attain by the recipient. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they had been 20 years
older. Ignoring all other features, this would comply with the
definition of plausibility if 40-year-old individuals were in fact more
credit-worthy on average than young adults. But of course this CE does
not qualify for providing actionable recourse to the applicant since
\emph{age} is not a (directly) mutable feature. Counterfactual training
aims to improve model explainability by aligning models with
counterfactuals that meet both desiderata: plausibility and
actionability. Formally, we define explainability as follows:

@@

We also find that CT can reduce sensitivity to immutable, protected
features and thus lead to less costly counterfactual outcomes as shown
in Figure~\ref{fig-poc}. In column three of Table~\ref{tbl-main}, we
impose mutability constraints on selected features and compute the
reduction in average costs of CEs associated with CT compared to the
baseline: for synthetic datasets, we always protect the first feature;
for all real-world tabular datasets we could identify and protect an
\emph{age} variable; for \emph{MNIST}, we protect the five top and
bottom rows of digits. Reductions in costs are overwhelmingly positive
and significant of up to nearly 60\% for \emph{GMSC}. While the
estimated cost reductions for \emph{Adult} and \emph{MNIST} are not
significant, Figure~\ref{fig-mnist} (columns 6-10) demonstrates that CT
does have the expected effect: sensitivity to protected features as per
integrated gradients is drastically reduced; details of this experiment
are reported in the supplementary appendix. In the case of \emph{Cred},
average costs increase, likely because any potential benefits from
protecting the \emph{age} are outweighed by the increase in costs
required for greater plausibility.

@@

\caption{\label{tbl-main}Key performance metrics and bootstrapped
\emph{two} standard errors for all datasets. \textbf{Plausibility}
(columns 1-2): percentage reduction in implausibility for \(IP\) and
\(IP^*\), respectively; \textbf{Actionability} (column 3): percentage
reduction in costs with protected features. \textbf{Accuracy} (columns
4-7): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
and the baseline (BL). Counterfactual outcomes in columns 1-3 are
aggregated across bootstrap samples and varying degrees of the energy
penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time.
Standard errors for accuracy are bootstrapped from the test set.}

>>>>

\iffalse
%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \title{My Publication Title --- Multiple Authors}
\author {
    % Authors
    First Author Name\textsuperscript{\rm 1},
    Second Author Name\textsuperscript{\rm 2},
    Third Author Name\textsuperscript{\rm 1}
}
\affiliations {
    % Affiliations
    \textsuperscript{\rm 1}Affiliation 1\\
    \textsuperscript{\rm 2}Affiliation 2\\
    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
}
\fi
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\usepackage{placeins}
\usepackage{siunitx}
\sisetup{uncertainty-mode = separate}

@@

Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
training regime explicitly geared towards improving the explanatory
capacity of models that, in high-level terms, we define as the extent to
which valid counterfactual explanations derived for a model can be
deemed plausible with respect to the underlying data and global
actionability constraints (we refine this notion in Def.
\ref{def-explainability}). To the best of our knowledge, Counterfactual
Training represents the first attempt to achieve more explainable models
by employing counterfactual explanations already in the training phase.

@@

In the context of counterfactual explanations, plausibility has broadly
been defined as the degree to which generated CEs comply with the
underlying data-generating process
\citep{altmeyer2024faithful, guidotti2022counterfactual, poyiadzi2020face}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. An AR recommendations must also be actionable, i.e.,
possible to attain by the recipient. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they had been 20 years
older. Ignoring all other features, this would comply with the
definition of plausibility if 40-year-old individuals were in fact more
credit-worthy on average than young adults. But to the applicant, this
CE does not provide actionable recourse, as age is not a (directly)
mutable feature. Counterfactual training aims to improve model
explainability by aligning models with counterfactuals that meet both
desiderata: plausibility and actionability. Formally, we define
explainability as follows:

@@

We also find that CT can reduce sensitivity to immutable, protected
features and thus lead to less costly counterfactual outcomes as shown
in Figure~\ref{fig-poc}. In column three of Table~\ref{tbl-main}, we
impose mutability constraints on selected features and compute the
reduction in average costs of CEs associated with CT compared to the
baseline: for synthetic datasets, we always protect the first feature;
for all real-world tabular datasets we could identify and protect an
\emph{age} variable; for \emph{MNIST}, we protect the five upper and
lower pixel rows of the full image. Reductions in costs are
overwhelmingly positive and significant of up to nearly 60\% for
\emph{GMSC}. While the estimated cost reductions for \emph{Adult} and
\emph{MNIST} are not significant, Figure~\ref{fig-mnist} (columns 6-10)
demonstrates that CT does have the expected effect: sensitivity to
protected features as per integrated gradients is drastically reduced;
details of this experiment are reported in the supplementary appendix.
In the case of \emph{Cred}, average costs increase, likely because any
potential benefits from protecting the \emph{age} are outweighed by the
increase in costs required for greater plausibility.

@@

\caption{\label{tbl-main}Key performance metrics and bootstrapped
\emph{two} standard errors for all datasets. \textbf{Plausibility}
(columns 1-2): percentage reduction in implausibility for \(\text{IP}\)
and \(\text{IP}^*\), respectively; \textbf{Actionability} (column 3):
percentage reduction in costs with protected features. \textbf{Accuracy}
(columns 4-7): test accuracies and robust accuracies (\(\text{Acc}^*\))
for CT and the baseline (BL). Counterfactual outcomes in columns 1-3 are
aggregated across bootstrap samples and varying degrees of the energy
penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time.
Standard errors for accuracy are bootstrapped from the test set.}

### 2025-07-14 paper/sections/experiments.qmd

We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as shown in @fig-poc. In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of CEs associated with CT compared to the baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom rows of digits. Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*. While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

@@

Key performance metrics and bootstrapped *two* standard errors for all datasets. **Plausibility** (columns 1-2): percentage reduction in implausibility for $IP$ and $IP^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features. **Accuracy** (columns 4-7): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL). Counterfactual outcomes in columns 1-3 are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time. Standard errors for accuracy are bootstrapped from the test set.
:::

>>>>

We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as shown in @fig-poc. In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of CEs associated with CT compared to the baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image. Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*. While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

@@

Key performance metrics and bootstrapped *two* standard errors for all datasets. **Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features. **Accuracy** (columns 4-7): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL). Counterfactual outcomes in columns 1-3 are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time. Standard errors for accuracy are bootstrapped from the test set.
:::
